Iterative Monomorphisation

### 2024-02-20 iterative_monomorphisation.tex
edits with Tanguy

\begin{definition}%[Type]
A \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\), \(\ty{map}(\ty{int},\ty{string})\)).
If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
\end{definition}

\begin{definition}%[Monomorphism and polymorphism]
A type is \emph{monomorphic} if it contains no type variables; otherwise it is \emph{polymorphic}.
\end{definition}

@@

\begin{definition}
A (type) \emph{substitution} is a partial function mapping a finite number of type variables to corresponding types. Substitutions are written as
$\sigma = \{\alpha_1\mapsto\tau_1, \dots, \alpha_n\mapsto\tau_n\}$. They are assumed to be lifted to formulae; thus, $\sigma(\varphi)$ yields the variant of $\varphi$ in which each $\alpha_i$ is replaced by $\tau_i$.
\end{definition}

\begin{definition}
Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_n\mapsto\tau_n\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_m\mapsto\upsilon_m\}\) are \emph{compatible} if for all \(i, j\), \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\). 
\end{definition}

\begin{definition}
Given two types \(\tau, \upsilon\), \emph{matching} \(\upsilon\) against \(\tau\) will either fail or yield a substitution \(\sigma\) such that \(\sigma\upsilon = \tau\).
\end{definition}

@@

We consider an initial problem \(\Phi\) to be a set of formulae \(\varphi_1, \dots, \varphi_n\).

\subsection{Collection}
The initial phase of each iteration consists in collecting the elements of the maps \(M\) and \(P\).

Given a type \(n\)-ary symbol \(f\) in \(\Phi\). \(M(f)\) is the set of all \(n\)-tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\). For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) (with \(\tau_1, \tau_2\) and \( \tau_3\) monomorphic) occurs in \(\Phi\) then, \((\tau_1, \tau_2, \tau_3) \in M(f) \).

Given a formula \(\varphi \in \Phi\) and a type \(n\)-ary symbol \(f\) in \(\varphi\). \(P(\varphi, f)\) is the set of all remaining \(n\)-tuples of type arguments to which \(f\) is applied in \(\varphi\) which are not in \(M(f)\). Meaning that at least one the type arguments is polymorphic. For example, if \(f\langle \tau_1, \tau_2, \tau_3\rangle\) (with at least one of \(\tau_1, \tau_2\) or \( \tau_3\) polymorphic) occurs in \(\varphi\) then, \((\tau_1, \tau_2, \tau_3) \in P(\varphi, f) \).

It is important to parametrise \(P\) with both \(\varphi\) and \(f\) because type variables are quantified at the formula level. This is not necessary for \(M\) since all the types it contains are monomorphic.

\subsection{Iteration}
Each iteration of the algorithm follows these steps:
\begin{itemize}

>>>>

\begin{definition}%[Type]
A (\emph{polymorphic}) \emph{type} \(\tau\) can be a type variable (e.g.\ \(\alpha\)) or
the application of an \(n\)-ary type constructor to \(n\) types (e.g.\ \(\ty{list}(\alpha)\), \(\ty{map}(\ty{int},\ty{string})\)).
If $n = 0$, we omit the parentheses (e.g.\ \(\ty{int}\)).
\end{definition}

\begin{definition}%[Monomorphism and polymorphism]
A type is \emph{monomorphic} if it contains no type variables.
\end{definition}

@@

\begin{definition}
A (\emph{type}) \emph{substitution} is a partial function mapping a finite number of type variables to corresponding types. Substitutions are written as
$\sigma = \{\alpha_1\mapsto\tau_1, \dots, \alpha_n\mapsto\tau_n\}$. They are assumed to be lifted to formulae; thus, $\sigma(\varphi)$ yields the variant of $\varphi$ in which each $\alpha_i$ is replaced by $\tau_i$.
\end{definition}

\begin{definition}
Two substitutions \(\{\alpha_1 \mapsto \tau_1, \dots, \alpha_m\mapsto\tau_m\}\) and \(\{\beta_1 \mapsto \upsilon_1, \dots, \beta_n\mapsto\upsilon_n\}\) are \emph{compatible} if \(\alpha_i = \beta_j\) implies \(\tau_i = \upsilon_j\) for all \(i, j\).
\end{definition}

\begin{definition}
Given two types \(\tau, \upsilon\), \emph{matching} \(\upsilon\) against \(\tau\) will either fail or yield a substitution \(\sigma\) such that \(\sigma(\upsilon) = \tau\).
\end{definition}

@@

We start with a high level description of the iterative monomorphisation algorithm. The algorithm takes a polymorphic problem as argument and returns a monomorphic problem. It operates by applying an arbitrary number of iterations. Each iteration takes a polymorphic problem as argument and returns a problem with additional partially instantiated formulae. A single iteration consists of a collection phase (Section~\ref{ssec:collection}) and a processing phase (Section~\ref{ssec:processing}).

\subsection{Collection}
\label{ssec:collection}

The initial phase of each iteration consists of computing two maps, \(M\) and \(P\), from the input problem~$\Phi$:

\begin{itemize}
\item Given a symbol \(f\) occurring in \(\Phi\), the set \(M(f)\) consists of all tuples of monomorphic type arguments to which \(f\) is applied in \(\Phi\). For example, if \(\sym{foldl}\langle \ty{nat}, \ty{int}\rangle\) occurs in \(\Phi\), then \((\ty{nat}, \ty{int}) \in M(\sym{foldl}) \).

\item Given a formula \(\varphi \in \Phi\) and a symbol \(f\) occurring in \(\varphi\), the set \(P(\varphi, f)\) consists of all tuples of type arguments to which \(f\) is applied in \(\varphi\) and which contains a type variable. For example, if \(\sym{foldl}\langle \ty{nat}, \ty{list}(\alpha)\rangle\) occurs in \(\varphi\), then \((\ty{nat}, \ty{list}(\alpha)) \in P(\varphi, \sym{foldl}) \).
\end{itemize}

It is important to parametrise \(P\) with \(\varphi\) because type variables are quantified at the formula level. The formula indicates the scope of type variables. This is not necessary for \(M\) since all the types it contains are monomorphic.

\subsection{Processing}
\label{ssec:processing}

Each iteration of the algorithm follows these steps:
\begin{itemize}
